---
title: "Bike Share Report"
output: html_document
---

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(plotly)

# Read in csv files
ny <- read.csv('new_york_city.csv')
wash <- read.csv('washington.csv')
chi <- read.csv('chicago.csv')
```

```{r message=FALSE, warning=FALSE}
# Displays first rows of each dataframe
head(ny)

head(wash)

head(chi)
```

```{r}
# Data manipulation 

calculate_age <- function(df, birth_year, start_time) {
  #' Calculates user age by subtracting birth year from start year

  # Converts Start.Time to Date datatype
  df[[start_time]] <- as.Date(df[[start_time]])
  
  # Extracts year from Start.Time
  start_year <- year(df[[start_time]])
  
  # Calculates difference in years
  age_difference <- start_year - df[[birth_year]]

  # Returns age
  return(age_difference)
}

# Creating new column for age and applying function
ny$age <- calculate_age(ny, "Birth.Year", "Start.Time")
chi$age <- calculate_age(chi, "Birth.Year", "Start.Time")

# Creating new column for city name
ny$city <- "New York"
chi$city <- "Chicago"
wash$city <- "Washington"

# Combining ny and chi dataframes into one dataframe
ny_chi <- rbind(ny, chi)

# Removing dissimilar columns compared to wash
ny_trunc <- ny %>%
  select(-one_of("Gender", "Birth.Year", "age"))

chi_trunc <- chi %>%
  select(-one_of("Gender", "Birth.Year", "age"))

# Combining ny, chi, and wash dataframes into one dataframe
all_city <- rbind(ny_trunc, chi_trunc, wash)
```

### How do subscriber ages compare by gender in Chicago versus New York?

<br></br>

```{r warning=FALSE, message=FALSE}
subscriber_age <- ny_chi %>%
  # Filters for Subscribers, non-null age, and non-null gender
  filter(User.Type == "Subscriber" & !(is.na(age)) & Gender=="Male" | Gender=="Female") %>%
  group_by(city, Gender, age) %>%
  summarise(subscriber_count = n())

# Finds highest subscriber count for females in New York
max_f_ny <- subscriber_age %>%
  filter(city == "New York", Gender == "Female") %>%
  arrange(desc(subscriber_count)) %>%
  slice(1)

max_f_ny

# Finds highest subscriber count for males in New York
max_m_ny <- subscriber_age %>%
  filter(city == "New York", Gender == "Male") %>%
  arrange(desc(subscriber_count)) %>%
  slice(1)

max_m_ny

# Finds highest subscriber count for females in Chicago
max_f_chi <- subscriber_age %>%
  filter(city == "Chicago", Gender == "Female") %>%
  arrange(desc(subscriber_count)) %>%
  slice(1)

max_f_chi

# Finds highest subscriber count for males in Chicago
max_m_chi <- subscriber_age %>%
  filter(city == "Chicago", Gender == "Male") %>%
  arrange(desc(subscriber_count)) %>%
  slice(1)

max_m_chi

# Finds count for females and males in each city
f_m_count <- ny_chi %>%
  filter(User.Type == "Subscriber" & !(is.na(age)) & Gender=="Male" | Gender=="Female") %>%
  group_by(city, Gender) %>%
  summarise(count = n())

f_m_count

graph1 <- ny_chi %>%
  # Filters for Subscribers, non-null age, and non-null gender
  filter(User.Type == "Subscriber" & !(is.na(age)) & Gender=="Male" | Gender=="Female") %>%
  # Creates histogram with labels
  ggplot(aes(x = age)) + 
  geom_histogram(binwidth=5, color="black", fill="lightblue") + 
  labs(title="Distribution of Age by Gender in New York vs Chicago", x="Age", y="Count") + 
  # facets by gender and city
  facet_grid(city~Gender) +
  theme_minimal() + 
  scale_x_continuous(breaks = seq(0, 100, by=10))

# Converts to a plotly graph for tooltip functionality
ggplotly(graph1)
```

#### Summary 

The histograms display the distribution of age faceted by gender for New York and Chicago. 
While the counts for Chicago are lower than the counts for New York, it can be seen 
that for both cities, there are more men than women riders. In New York, there are 36,625 men
compared to 12,159 women. In Chicago, there are 5,159 men and 1,723 women. Both cities' histograms are
also skewed right for both men and women. It appears that most riders are between people
in their early twenties to mid-thirties, regardless of city. The most frequent age for 
female subscribers in New York is 28 while it is 32 for male subscribers. The most 
frequent age for female subscribers in Chicago is 25 while it is 28 for males in 
Chicago.


### How does the average trip duration compare by user type across different cities?

<br></br>

```{r warning=FALSE, message=FALSE}
# Creates new column for converting trip duration from seconds to minutes
all_city$Trip.Duration.Mins <- all_city$Trip.Duration / 60

avg_trip <- all_city %>%
  # Filters for non-null user types
  filter(User.Type == "Subscriber" | User.Type == "Customer") %>%
  group_by(city, User.Type) %>%
  # Calculating average trip duration by city and user type
  summarise(Average = mean(Trip.Duration.Mins, na.rm=T))

avg_trip

graph2 <- avg_trip %>%
  # Creates grouped bar chart with labels
  ggplot(aes(x=city, y=Average, fill=User.Type)) + 
  geom_col(position = "dodge") + 
  labs(title="Average Trip Duration by User Type by City", 
       x="City", 
       y="Trip Duration (Minutes)", 
       fill="User Type") +
  theme_minimal() + 
  scale_fill_brewer(palette="Paired")

# Converts to a plotly graph for tooltip functionality
ggplotly(graph2)

```
<br></br>

#### Summary

The grouped bar chart shows that across all cities in the data, customers take
longer average trips than subscribers. Customers in Washington seem to take the 
longest trips on average, followed by New York, and then Chicago at 43.9 minutes, 
36.6 minutes, and 32.2 minutes, respectively. The subscribers across all cities 
take similar trip duration on average at 11.4 minutes for Chicago, 
12.6 minutes for New York and 12.2 minutes for Washington. This could be due to 
one-time customers tending to be tourists and subscribers tending to be locals 
using bike sharing for commuting to work, however this cannot be proven or denied
based on the dataset.


### How do daily and hourly usage rates compare by city?

<br></br>

#### New York

```{r message=FALSE, warning=FALSE}
extract_hour_day <- function(df, start_time) {
  #' Extracts the hour and day of the week from Start.Time and order the days of the week
  
  # Convert Start.Time to POSIXct datetime format
  df[[start_time]] <- as.POSIXct(df[[start_time]], format = "%Y-%m-%d %H:%M:%S")
  
  # Extract hour and day of the week from Start.Time
  df$hour <- format(df[[start_time]], "%H")
  df$day_of_week <- format(df[[start_time]], "%A")
  
  # Order days of the week
  df$day_of_week <- factor(df$day_of_week, levels=c("Sunday", "Saturday", "Friday", "Thursday", "Wednesday",
                                                    "Tuesday", "Monday"))
  return(df)
}

usage_df <- function(df) {
  #' Filters, groups by day of the week and hour, and finds trip count for the df

  # Filters and summarises dataframe
  usage_detail <- df %>%
    filter(!(is.na(day_of_week)) & !(is.na(hour))) %>%
    group_by(day_of_week, hour) %>%
    summarise(trip_count = n(), .groups = 'drop')
  
  return(usage_detail)
}

plot_heatmap <- function(df) {
  #' Plots the usage detail df in a heatmap
  
   plot <- ggplot(df, aes(x = hour, y = day_of_week, fill = trip_count)) +
    # Creates heatmap with labels
    geom_tile() + 
    scale_fill_gradient(low = "lightblue", high = "darkblue") +
    theme_minimal() +
    labs(title = "Bike Usage by Hour and Day of the Week",
         x = "Hour of the Day",
         y = "Day of the Week",
         fill = "Trip Count")
   
   return(plot)
}

# Extracts hour and day of week from New York dataframe
hour_day_ny <- extract_hour_day(ny, "Start.Time")

# Returns usage dataframe
detail_ny <- usage_df(hour_day_ny)

# Finds top 5 highest usage times
usage_summary <- detail_ny %>%
  arrange(desc(trip_count)) %>%
  slice(c(1, n()))

# Prints top 5 usage times
usage_summary

# Returns heatmap for New York dataframe
plot_heatmap(detail_ny)
```


#### Chicago

<br></br>

```{r}
# Extracts hour and day of week from Chicago dataframe
hour_day_chi <- extract_hour_day(chi, "Start.Time")

# Returns usage dataframe
detail_chi <- usage_df(hour_day_chi)

# Finds top 5 highest usage times
usage_summary2 <- detail_chi %>%
  arrange(desc(trip_count)) %>%
  slice(c(1, n()))

# Prints top 5 usage times
usage_summary2

# Returns heatmap for Chicago dataframe
plot_heatmap(detail_chi)
```

#### Washington D.C.

<br></br>

```{r}
# Extracts hour and day of week from Washington dataframe
hour_day_wash <- extract_hour_day(wash, "Start.Time")

# Returns usage dataframe
detail_wash <- usage_df(hour_day_wash)

# Finds top 5 highest usage times
usage_summary3 <- detail_wash %>%
  arrange(desc(trip_count)) %>%
  slice(c(1, n()))

# Prints top 5 usage times
usage_summary3

# Returns heatmap for Washington dataframe
plot_heatmap(detail_wash)
```

<br></br>

#### Summary

The heatmaps show different levels of usage throughout each hour of the day
by day of the week for each city. The heatmap for New York shows that the 
days and times with the heaviest usage are on weekdays between 7am - 9am and 
5pm - 7pm. Weekends during the day also seem to be popular in New York. The day 
and time with the highest usage in New York is Wednesday at 5pm while the lowest usage 
is Tuesday at 3am.

This pattern is also mirrored pretty similarly in Chicago where the heaviest 
usage times are weekdays from around 7am - 8am and around 4pm - 6pm. Weekends
during the day also seem to be popular. The day and time with the highest usage 
in Chicago is Thursday at 5pm and the lowest usage is Wednesday at 2 am. The 
weekday pattern could be due to professionals utilizing bike sharing to commute 
to and from work. The weekend patterns could be due to riders using the bikes for leisure.

Interestingly, in Washington D.C the heaviest usage times are on weekdays 
from around 5am - 8am, however this is not mirrored in the evening as expected. 
For weekends, it seems that the late night/ early morning time is the most popular.
The day and time with the highest usage is Wednesday at 8am and the lowest usage is
Thursday at 12am. 

